---
title: "HLC Day 3 Homework 3"
author: "Brooke Felsheim, Elena Kharitonova, Ellen Risemberg"
date: "6/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 3

During class today, we went over how to load your own data into R. For today's
homework, we will practice saving a data frame as a new file.

## Writing data frames to a file

In homework 2, we used a subset of data from the `gapminder` library to generate
a plot. Let's try saving this subsetted data frame to its own file in your local
filesystem. Start by loading the gapminder library and creating the subsetted
data frame by running the code below.

```{r}
library(gapminder)
homeworkData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
```

We will save two different versions of this file. First, we will save this as 
a tab-separated text file. We can do this with the `write.table` function, which
is similar to the `read.table` function, but instead of reading in a table, it
writes a table to a file.

### Write to .txt file

Look at the help page for this function by typing `?write.table` in your R console.

You'll see that the first argument (`x`) of `write.table` expects a matrix or data frame.
The second argument (`file`) is where you can specify the file path and name. The other arguments change different characteristics about the way the file is saved. We may
want to override the default arguments for these.

NOTE: before saving a new file, check where your current working directory is with
`getwd()`. This is where the file will be saved if you do not specify a full path.
You can change this to a different directory using `setwd()` as we discussed in class,
or navigating to the folder you would like to use in the bottom right corner of your 
R Studio panel and selecting "More" -> "Set as Working Directory". In an R Markdown
file, the working directory defaults to the directory that the .Rmd file is saved in.

```{r}
getwd()
setwd(dir = ) # set working directory to another location if desired
```

First, let's try writing to a tab separated text file using the default parameters.

```{r}
write.table(x = homeworkData, file = "homework_3_data.txt", sep = "\t")
```

If you open the saved text file, you'll see that the row numbers of the data frame
were saved. You'll also see that all character values are output with quotes
around them. If you don't want the row values and/or these quotes, you can 
overwrite some of the default arguments in R. Hint: see `row.names` and `quote`
in the `write.table` help page.

Try saving the file again to "homework_3_data_2.txt", but without row names 
and quotes in the R chunk below.

```{r}

```

### Write to .csv file

Now try writing the homework data to a .csv file called "homework_3_data.csv". Try doing this yourself using the `write.table` function and setting `sep = ","` in the R chunk
below. Try removing the row names and the quotes as you did for the .txt file. 

```{r}

```

Note: you can also write as a .csv file using `write.csv` function. Try writing the
data again to a csv file called "homework_3_data_2.csv", but using `write.csv` in 
the R chunk below. 

```{r}

```


