---
title: "HLC_Homework6"
author: "Ellen Risemberg"
date: "6/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exercise 1

Use `genPrimer` from functions day 1 to generate a list of 20 primers that are between 20-30 base pairs in length.

`genPrimer` function:

```{r}
library(stringi) # Required for stri_rand_strings function

genPrimer <- function(minlen, maxlen, nprimers, bases = '[ACTG]') {
  
  lengths <- lengths1 <- sample(minlen:maxlen, nprimers,replace=TRUE)
  res <- stri_rand_strings(nprimers, lengths, bases)
  return(res)
  
}
```

Create primers vector: 

```{r}
primers <- genPrimer(20, 30, 20)
```

Write a for loop that prints "primer is between 20 and 24 bases" if the primer has 20 or more bases and 24 or fewer bases. 

```{r}
for(primer in primers) {
  if(nchar(primer) > 19 & nchar(primer) < 25) print("primer is between 20 and 24 bases")
}
```

#### Exercise 2

Write a function that takes as input a gapminder dataset from 2002 and a specified continent, and prints out the list of countries.

```{r}
library(dplyr)
library(gapminder)

gap2002 <- filter(gapminder, year==2002)

printCty <- function(data, cont)
{

  for (i in 1:length(data$continent))
  {
    if (data$continent[i] == cont) 
    {
      print(as.character(data$country[i]))
    }
  }
}

printCty(gap2002, "Americas")
printCty(gap2002, "Africa")
printCty(gap2002, "Asia")
```

