---
title: "HLC_Homework5"
author: "Ellen Risemberg"
date: "6/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 1

Write a function that returns the length and sum of a given vector of numbers. How can you work around the fact that `return` can only be used once in a function?

**Solution**:

```{r}
```

### Exercise 2

The following function generates random primers. Use it to create a list of 50 primers between 40-50 base pairs:

```{r}
library(stringi) # needed for function stri_rand_strings

genPrimer <- function(minlen, maxlen, nprimers, bases = '[ACTG]') {

  lengths <- lengths1 <- sample(minlen:maxlen, nprimers,replace=TRUE)
  res <- stri_rand_strings(nprimers, lengths, bases)
  return(res)
  
}
```

Then, write a function that counts the number of G's in each of your randomly generated primers from above, and returns three items: the primer sequence, the primer length, and the number of G's in that primer.  

*Hint:* Use the function `str_count` in the library `stringr` to count the G's.

**Solution**:

```{r}
```


