---
title: "HLC_Homework6"
author: "Ellen Risemberg"
date: "6/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exercise 1

Use `genPrimer` from functions day 1 to generate a list of 20 primers that are between 20-30 base pairs in length.

`genPrimer` function:

```{r}
library(stringi) # Required for stri_rand_strings function

genPrimer <- function(minlen, maxlen, nprimers, bases = '[ACTG]') {
  
  lengths <- lengths1 <- sample(minlen:maxlen, nprimers,replace=TRUE)
  res <- stri_rand_strings(nprimers, lengths, bases)
  return(res)
  
}
```

Create primers vector: 

```{r}
primers <- genPrimer(20, 30, 20)
```

Write a for loop that prints "primer is between 20 and 24 bases" if the primer has 20 or more bases and 24 or fewer bases. 

```{r}
for(primer in primers) {
  if(nchar(primer) > 19 & nchar(primer) < 25) print("primer is between 20 and 24 bases")
}
```

#### Exercise 2

Make a vector of the days of the week:

```{r}
days <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
```

Use sapply to create a vector that contains abbreviations for the days of the week using days. Use an if/else statement to keep just the first character of each day of the week that starts with a unique letter (Monday, Wednesday, Friday), but keep two characters for days that start with the same letter as another day (Tuesday, Thursday, Saturday, and Sunday).

Hint: use the function substr to extract a subset of the characters from a character string.

```{r}
abbr <- sapply(days, function(x) 
  if(substr(x, 1, 1) == "T" | substr(x, 1, 1) == "F") {substr(x, 1, 2)} # if the first character is T or F, extract first two characters
               else {substr(x, 1, 1)} # everything else, extract first character
  )
```

