---
title: "Functions and control flow"
author: "Ellen Risemberg"
date: "June 21, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

So far we've been using a lot of functions that exist within R. Now we'll learn how to build our own function.


```{r eval=FALSE}
myfunction <- function(arg1, arg2, ...) {
  
  #do some stuff
  return(object)
  
}
```

Make a simple function that just returns the argument that it is given.

```{r}
first_function <- function (x) {
  
  return(x)
  
}
```

Now if you pass a value to `first_function` it should return that value:

```{r}
first_function(9)
```

A function that only returns the given value is not useful, but it does illustrate how you can add functions to your environment. Now make a function that squares a value and then adds one to it.

```{r}
second_function <- function(x) {
  ans <- x^2 + 1
  return(ans)
  
}

second_function(9)
```


#### Exercise 1

Write a function that converts a Fahrenheit temperature to Kelvin (remember that K = 5/9*(F - 32) + 273.15).

```{r}
f2k <- function(f) {
  k <- (5/9)*(f - 32) + 273.15
  return(k)
}

f2k(60)
```

#### Exercise 2

Suppose we wanted to look at the Gapminder dataset and calculate GDP. Write a function to do that, returning the dataset with the new GDP column

Tip: first, load `dplyr` and `gapminder`. 

```{r}
library(dplyr)
library(gapminder)
```

Solution:

```{r}
getGDP <- function(data){
  new_data <- mutate(data, GDP=gdpPercap*pop)
  return(new_data)
}

getGDP(gapminder)
```

Alternate solution:

```{r}
# Makes general function that computes GDP from 2 columns, which can be passed to mutate

getGDP2 <- function(data){
  gdp <- data$gdpPercap * data$pop
  return(gdp)
}

# Returns a vector of GDPs
getGDP2(gapminder)

# Inside mutate will capture the vector in a column named `GDP`
gapminder %>% 
  dplyr::mutate(GDP = getGDP2(.))

```


## Control flow 

Flow of control through any given function is implemented with various **control structures**. There are three basic types of control structures:
1. Sequential: the default mode. Sequential execution of code, one line after another. 
2. Selection: used for decisions and branching, such as choosing between two or more alternative paths. 
  * `if` statements
  * `if/else` statements
3. Repetition: used for looping, i.e. repeating a piece of code multiple times. 
  * `for` loops
  * `while` loops
  * `do/while` loops

### For loops 

For loops are used to iterate through elements in a vector, repeating the same block of code each time. They take the basic form:

```{r, eval=FALSE}
for(i in vector) {
  Do some operation to each element of the vector
}
```

Example:

```{r}
years <- c(2010,2011,2012,2013,2014,2015)

for (year in years){
  print(paste("The year is", year))
}
```

We can also store the output of the code block evaluated in each iteration of the for loop:

```{r}
output <- c() # create an empty vector

for (year in years){
  output <- c(output, paste("The year is", year)) # add to the empty vector each iteration
}

output
```

We can also do this another way by iterating through the indices with the `length` function. This is rather common in more traditional programming languages, such as MATLAB, C++, and Java.

```{r}
output <- c() # re-initialize empty vector

for (i in 1:length(years)){
  output[i] <- paste("The year is", years[i]) # add to the empty vector each iteration
}

output
```

**Note**: you may hear people say that you "shouldn't" use `for` loops in R (vs other languages). In general this is not true and you will probably use `for` loops frequently; however, depending on the problem there is sometimes a more efficient option:
1. `apply` family of functions (discussed later)
2. Vector-based methods, aka *vectorization* (discussed now) 

A simple example comparing for loop vs vectorization in performing vector addition.

Create two vectors of random data:

```{r}
a <- runif(n = 100, min = 0, max = 1)
b <- runif(n = 100, min = 0, max = 1)
```

Using for loop:

```{r}
sum <- vector()

for (i in 1:length(a)){
  sum[i] = a[i] + b[i]
}

sum
```

Using a vector-based method:

```{r}
sum <- a + b

sum
```


#### Exercise 3

Use a for loop to print out all the countries in the Americas from Gapminder (hint: use `filter` to just extract the American countries; also, how can you avoid repeats?)

```{r}
us_Gapminder <- filter(gapminder, continent == "Americas" & year == 2002)

for (i in 1:length(us_Gapminder$country)){
  print(us_Gapminder$country[i])
}
```

The levels tacked on make our output a little hard to read. We can suppress those by using the `as.character` function.

```{r}
for (i in 1:length(us_Gapminder$country)){
  print(as.character(us_Gapminder$country[i]))
}
```

If you want to keep the output of a for loop, it can be more efficient to use a function from the `apply` family. We'll go over this later. 


#### Exercise 4

Use the following vector `years` and the function `printYear` inside a for loop to print a new line for each year in `years`.

```{r}
years <- c(2010,2011,2012,2013,2014,2015)

printYear <- function(year) {
  print(paste("The year is", year))
}
```

**Solution**:

```{r}
for (year in years){
  printYear(year)
}
```


### If and if/else statements

Like `for` loops, `if` and `if/else` statements are an important part of programming in all languages. They are used for decision making. 
* `if` statements are used to determine whether or not a code block is executed. 
* `if/else` statements are used to choose between 2 or more alternate paths. 

The code within an `if` or `if/else` statement is executed if the specified condition is TRUE. 

In R, you create an `if/else` statement with the following syntax:

```{r, eval=FALSE}
if (logical statment) { do something } else  { do something different } 
```

For example:

```{r}
if (1 > 2){ 
  print("TRUE")
} else {
  print("FALSE")
}
```

The condition following the `if` statement is a conditional that evaluates to a `boolean` value (either TRUE or FALSE). So you could equivalently use this code: 

```{r}
print(1 > 2)
```

Not all `if/else` statements have to have an `else` clause. By default, if the condition evaluates to FALSE and no `else` statement is given, R does nothing. This is just an `if` statement. 

```{r}
x <- 1000
if (x > 100) print("'x' is greater than 100")
```

```{r}
x <- 10
if (x > 100) print("'x' is greater than 100")
```

If/else statements can also be nested:

```{r}
y <- 25

if (y < 100) {
  if (y > 10) {
    print("10 < y < 100")
  } else {
    print("y < 10 or y > 100")
  }
} else {
  print("y < 10 or y > 100")
}
```


#### Exercise 5

Without using `filter`, loop through the entire Gapminder dataset, and just print out countries in the American continent. For simplicity and to avoid repeats, just use data from the year 2002: `gap2002 = filter(gapminder, year==2002)`

```{r}
gap2002 <- filter(gapminder, year==2002)

for (i in 1:length(gap2002$continent))
{
  if (gap2002$continent[i] == "Americas") 
  {
    print(as.character(gap2002$country[i]))
  }
}
```

#### Bonus

Write a function that takes as input a gapminder dataset from 2002 and a specified continent, and prints out the list of countries.

```{r}
printCty <- function(data, cont)
{

  for (i in 1:length(data$continent))
  {
    if (data$continent[i] == cont) 
    {
      print(as.character(data$country[i]))
    }
  }
}

printCty(gap2002, "Americas")
printCty(gap2002, "Africa")
printCty(gap2002, "Asia")
```


### Apply family functions

Apply family functions similarly perform operations iteratively on slices of data (rows in a matrix, elements in a list). For many operations, the apply family can be more efficient than a for loop. You can read on about apply functions at [Data Camp](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family#family).

We can use `sapply`, which applies a function over a list or vector and returns the output as a *vector*, to get output as the above `for` loop:

```{r}
years <- c(2010,2011,2012,2013,2014,2015)
sapply(years, function(x) paste("The year is", x))
```

We can use `lapply`, which applies a function over a list or vector and returns the output as a *list*.

```{r}
years <- c(2010,2011,2012,2013,2014,2015)
lapply(years, function(x) paste("The year is", x))
```

#### Exercise 6

Repeat the Gapminder exercise (print out all countries in the Americas) using `sapply` and `lapply`

```{r}
us_Gapminder <- filter(gapminder, continent == "Americas" & year == 2002)

sapply(as.character(us_Gapminder$country), function(x) x)
lapply(as.character(us_Gapminder$country), function(x) x)
```


## A note on scope 

Scope is a very important concept in programming. It indicates a variable's "lifetime" in the program and defines **where variables can be accessed or referenced**.

To illustrate this, consider this function which adds 5 to the argument `x`. 

```{r}
add5 <- function(x){
  x <- x+5
  print(x)
}
```

We can set a variable and call the function on that variable: 

```{r}
num <- 6

add5(num)
```

What do you think the value of `num` will be at this point?

Note that we can also use the variable name `x` outside of the function, and this will be a separate variable. 

```{r}
x <- 7 
add5(x)
```

Read [Scoping Rules of R](https://bookdown.org/rdpeng/rprogdatascience/scoping-rules-of-r.html) for a more detailed explanation. 
