---
title: "HLC_Homework5"
author: "Ellen Risemberg"
date: "6/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 1

Write a function that returns the length and sum of a given vector of numbers. How can you work around the fact that `return` can only be used once in a function?

**Solution**:

```{r}
v <- rnorm(50, mean = 5) # create a vector of length 50 from random samples of a normal dist. with mean 5

myFun <- function(vector) {
  length <- length(vector)
  sum <- sum(vector)
  results <- list(length = length, sum = sum)
  return(results)
}

myFun(v)
```

### Exercise 2

The following function generates random primers. Use it to create a list of 50 primers between 40-50 base pairs:

```{r}
library(stringi) # needed for function stri_rand_strings
library(stringr)

genPrimer <- function(minlen, maxlen, nprimers, bases = '[ACTG]') {

  lengths <- lengths1 <- sample(minlen:maxlen, nprimers,replace=TRUE)
  res <- stri_rand_strings(nprimers, lengths, bases)
  return(res)
  
}
```

**Solution** to part 1:

```{r}
primers <- genPrimer(40, 50, 50)
```

Then, write a function that counts the number of G's in each of your randomly generated primers from above, and returns three items: the primer sequence, the primer length, and the number of G's in that primer.  

*Hint:* Use the function `str_count` in the library `stringr` to count the G's.

**Solution** to part 2:

```{r}
primerStats <- function(primerlist) {
  lengths <- nchar(primerlist)
  nG <- str_count(primerlist, "G")
  results <- data.frame(sequence = primerlist, length = lengths, numberG = nG)
  return(results)
}
```


