---
title: "Homework 2"
author: "Elena Kharitonova, Brooke Felsheim, Ellen Risemberg"
date: "6/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# HLC Homework 2: Plotting

For the second homework, your goal will to replicate the below plot to the best
of your ability using the `gapminder` data frame and `ggplot2`.

```{r echo = FALSE, message = FALSE}
library(gapminder)
library(ggplot2)
homeworkData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed") + 
ylab("Life expectancy") + 
  xlab("GDP per capita") + 
  labs(color = "Continent", size = "Population") + 
  ggtitle("Life expectancy increases independently of GDP per cap") +
  facet_wrap(vars(year)) + 
  theme_bw() +
  theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```

This plot is more complicated than any of the examples we have shown during
class, so we will provide hints and walk through how to replicate it in steps.

In addition to the hints that will be given, make use of your resources from the 
HLC markdown documents we have gone over in class, [R for data science](https://r4ds.had.co.nz/data-visualisation.html), and Google. Remember
that you can access the R documentation for any function by typing `?` before
the name of the function in the R console, or by searching for the function in the 
search box of the "Help" tab of the bottom right corner of R Studio.

## Steps for creating the plot

### Step 1

Create your own new R markdown document (.Rmd file) or R script (.R file) to run 
the code for this homework response. If you are using a .Rmd file, remember that 
you must run R code in the gray code blocks. We went over how to create these
during class.

### Step 2

Load the `gapminder` and `ggplot2` package contents into R. Remember that to load
package contents into R, we use the `library` function.

### Step 3

Run the below code to subset of the Gapminder data for the years 1952 and 2007. 
This will be the data frame that you supply to `ggplot()` in later steps. We
will cover subsetting data frames in more detail next week.
 
```{r}
homeworkData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
```

### Step 4

We will be plotting life expectancy (`lifeExp`) vs GDP per capita (`gdpPercap`) 
for this data. Start by replicating the below plot, referencing the examples we 
covered during class as needed.

```{r echo=FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
```

### Step 5

Now, let's color the points by continent and scale the size of the points by
population. Again, reference the examples where we did this in class as needed. 

```{r echo=FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop))
```

### Step 6

Notice that many of the points are overlapping each other. If we would like to
see where points overlap, we can alter the transparency with the `alpha` parameter.
`alpha` is a value that ranges from 0 (fully transparent) to 1 (fully opaque).
In the `geom_point()` part of your code, supply `alpha = 0.5` outside of the
mapping `aes()` specification. Your plot should now look like the one below.

```{r echo=FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5)
```

### Step 7

In the above plot, you may have noticed that the GDP per capita data on the x-axis
is mostly concentrated towards 0, with some large data outliers. It might be
better to transform the y-axis into a log scale. Let's set a log10 scale x-axis
for our plot. Hint: the function `scale_x_log10()` may be useful here. 

Your plot should now look like this:

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10()
```

### Step 8

It looks like there is a linear trend between life expectancy and GDP per capita.
We can add a linear trendline to our plot to show this relationship. To do this,
we will want to make use of the `geom_smooth` function. You can read more about 
how to use `geom_smooth` to plot a linear trendline [here](https://www.statology.org/ggplot-trendline/) and by looking at the R
documentation for this function. Bonus: hide the shaded confidence region with
the `se` parameter, change the line color to black, and set the line type to 
be dashed.

Your plot will now look like this:

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed")
```

You may get a message in R that says: `geom_smooth() using formula 'y ~ x'`.
You can ignore this.

### Step 9

Notice that all of our labels by default are using the column name. We want to
make these more descriptive. Let's change the x-axis label to "GDP per capita" 
and the y-axis label to "Life expectancy". Let's also add a title that says "Life 
expectancy increases independently of GDP per cap". We discussed how to do this 
in class.

We also would like to change the legend labels to "Continent" and "Population". 
Hint: there are three different ways to do this described [here](https://r-charts.com/ggplot2/legend/#title).

Your plot should now look like this:

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed") + 
ylab("Life expectancy") + 
  xlab("GDP per capita") + 
  labs(color = "Continent", size = "Population") + 
  ggtitle("Life expectancy increases independently of GDP per cap")
```

### Step 10

If you remember from the beginning, we filtered the `gapminder` data set to only
contain data from the years 1952 and 2007. We want to see if the trend of life
expectancy by GDP per capita is different between these two years by plotting 
the data for each year side-by-side. We can do this by using `facets` to show
the different years. We haven't covered this in class, but try looking at the R
help page for the `facet_wrap` function to make your plot look like the one below:

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed") + 
ylab("Life expectancy") + 
  xlab("GDP per capita") + 
  labs(color = "Continent", size = "Population") + 
  ggtitle("Life expectancy increases independently of GDP per cap") +
  facet_wrap(vars(year))
```

### Step 11

`ggplot2` has a theme system which allows you to have control over the non-data 
elements of your plot such as ticks, panel strips, and backgrounds. Themes can 
help you make the plot aesthetically pleasing or match an existing style guide. 
By default, `ggplot2` uses a theme called `theme_grey()`. This theme gives us the
gray background we've been seeing in our plot. We can change this theme to any
one of the `ggplot2` inbuilt complete themes which controll all non-data display.

Try experimenting with the theme. Run `?theme_grey` in the R console to see a
help page with the complete list of theme options. 

To match the plot below, change the theme to `theme_bw()`.

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed") + 
ylab("Life expectancy") + 
  xlab("GDP per capita") + 
  labs(color = "Continent", size = "Population") + 
  ggtitle("Life expectancy increases independently of GDP per cap") +
  facet_wrap(vars(year)) + 
  theme_bw()
```

### Step 12

Finally, we can use the `theme()` function if we want to tweak components of our
existing theme. One common component that you may want to change is the text size. 
Add the below code to your ggplot to resize the axis text to 12, the axis title 
text to 14, and the facet title text to 14.

```
theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```

Your plot should now look like this:

```{r echo = FALSE, message = FALSE}
ggplot(homeworkData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, size = pop), alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = lm, se = FALSE, color = "black", lty = "dashed") + 
ylab("Life expectancy") + 
  xlab("GDP per capita") + 
  labs(color = "Continent", size = "Population") + 
  ggtitle("Life expectancy increases independently of GDP per cap") +
  facet_wrap(vars(year)) + 
  theme_bw() +
  theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```

Your plot should now match the one at the beginning of the document! If you
couldn't get everything to match perfectly, don't worry. We will go over this in
class on Tuesday.

