---
title: "Homework 2"
author: "Elena Kharitonova, Brooke Felsheim, Ellen Risemberg"
date: "6/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# HLC Homework 2: Plotting

To check your understanding of `ggplot` and topics covered in today's lesson,
replicate the plot below of data from the Gapminder dataset to the best of your
ability. Make use of your resources the HLC Day2 R markdown document, the 
`ggplot2` documentation, R for Data Science, and Google. Create a new R script 
or R markdown document of your to run the code for your answer.

First, load the `gapminder` and `ggplot` libraries into R
 
```{r}
library(gapminder)
library(ggplot2)
```

This plot uses the following subset of the Gapminder data:
 
```{r}
finalExampleData <- dplyr::filter(gapminder, year %in% c(1952, 2007))
```

### Hints:

This plot has:     

 - Facets to show the different years. Try looking at the help page for `facet_wrap`.
 - A linear trendline
 - A log10 scale x-axis
 - Altered transparency of the points
 - Renamed legend titles
 - Axis text resized to 12
 - Axis title text resized to 14
 - Facet titles resized to 14
 

```{r gapminder plot, fig.height= 4, fig.width= 9, echo = F, message = F}
ggplot(finalExampleData, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(size = pop, color = continent), alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "black", lty = 2) +
  facet_wrap(~year) + 
  scale_x_log10() + 
  ylab("Life Expectancy") + 
  xlab("GDP per capita") + 
  guides(color = guide_legend("Continent"), 
         size = guide_legend("Population")) +
  ggtitle("Life Expectancy Increases independently of GDP per cap") +
  theme(strip.text.x = element_text(size = 14),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))
```

